### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding richardson iteration matrix spectral radius function

\section{Derivation of the $P_N$ Equations}
\label{sec:pn_equations}
Next, we derive the $P_N$ equations, a simplified form of the general
transport equation where Legendre polynomials are used to expand the
angular flux and scattering cross section variables as a means of
capturing the angular structure of the solution. Before deriving this
form of the transport equation, we briefly discuss a few properties of
Legendre polynomials that we will find useful in the derivation.

@@

\section{Derivation of the $SP_N$ Equations}
\label{sec:spn_equations}
The $P_N$ equations give $N+1$ coupled first-order equations
capturing the spatial and angular-dependence of the solution. In
multiple dimensions, the equation set becomes large and coupled not
only through angular moments but also through the spatial
variables. As a simpler alternative to multidimensional $P_N$
solutions, Gelbard recognized in 1960 that the planar $P_N$ equations
could be simplified and applied an ad-hoc method to extend them to
multiple dimensions, yielding the $SP_N$ equations. These equations
are not only fewer in number, but also take on a diffusion-like form
while maintaining the angular character of the flux, making them
amenable to solutions with modern diffusion methods.

@@

moments and a set of differential equations:

@@

\section{A Note on Spatial Discretization and Matrix Symmetry}
\label{sec:spatial_discretization}
To this point, the formulation of the multigroup $SP_N$ equations
presented have discretized the transport equation in angle and energy
but have yet to consider the spatial component of phase space. We
don't explicitly consider it in this document but instead we will
briefly comment on possible means of discretization. Of primary
interest here is the discretization of the diffusive term in
Eq~(\ref{eq:spn_multigroup_system}) and the convective term on the
boundaries in Eq~(\ref{eq:spn_multigroup_bnd}). Many popular choices
for spatial discretization are available here and include finite
difference and finite element formulations. In the Denovo package at
ORNL, Evans has implemented a finite volume scheme for these
equations. Although arbitrary grids could be handled effectively
through a finite element scheme, the rectilinear grid used in Denovo
is easily discretized through the finite volume method in a
conservative form. This conservative form is ideal for the
diffusion-like equations in the domain given by
Eq~(\ref{eq:spn_multigroup_system}) and the effective boundary current
conditions given by Eq~(\ref{eq:spn_multigroup_bnd}) as the neutron
current is balanced from cell-to-cell, continuity of the flux is
preserved across cell/material boundaries, and boundary conditions are
naturally represented through cell-face currents. Furthermore, this
spatial discretization is a natural extension of the balance
principles used to arrive at the general transport equation. The
result of this finite volume discretization is one that is symmetric
for all equations in the domain with respect to the spatial components
of the solution.

Given the symmetry of the spatial discretization equations in the
domain given by monoenergetic Eq~(\ref{eq:spn_matrix}) form a
symmetric linear operator acting on the pseudo-moment vector to give:
\begin{equation}
  \mathbf{L}\mathbf{u}=\mathbf{Q}\:,
  \label{eq:matrix_system}
\end{equation}
where $\mathbf{L} = -\nabla \cdot \mathbf{D} \nabla +
\mathbf{A}$. Moving to the multigroup formulation in
Eq~(\ref{eq:spn_multigroup_system}) we have:
\begin{equation}
  \mathbb{L}\mathbb{U}=\mathbb{Q}\:,
  \label{eq:multigroup_matrix_system}
\end{equation}
where $\mathbb{L} = -\nabla \cdot \mathbb{D} \nabla +
\mathbb{A}$. Whether or not the resulting matrix is symmetric is
determined by the group scattering matrices, $\sigma_{sn}^{gg'}$, for
each of the moments. In general, the cross sections in these matrices
do not form a symmetric matrix and therefore the resulting linear
operator in Eq~(\ref{eq:multigroup_matrix_system}) will always be
asymmetric for multigroup problems. Solutions to this linear system
will then require techniques specifically formulated for asymmetric
systems.

>>>>

%%---------------------------------------------------------------------------%%
\section{Derivation of the $P_N$ Equations}
\label{sec:pn_equations}
Next, we derive the $P_N$ equations, a simplified form of the general
transport equation where Legendre polynomials are used to expand the
angular flux and scattering cross section variables as a means of
capturing the angular structure of the solution. Before deriving this
form of the transport equation, we briefly discuss a few properties of
Legendre polynomials that we will find useful in the derivation.

@@

%%---------------------------------------------------------------------------%%
\section{Derivation of the $SP_N$ Equations}
\label{sec:spn_equations}
The $P_N$ equations give $N+1$ coupled first-order equations
capturing the spatial and angular-dependence of the solution. In
multiple dimensions, the equation set becomes large and coupled not
only through angular moments but also through the spatial
variables. As a simpler alternative to multidimensional $P_N$
solutions, Gelbard recognized in 1960 that the planar $P_N$ equations
could be simplified and applied an ad-hoc method to extend them to
multiple dimensions, yielding the $SP_N$ equations. These equations
are not only fewer in number, but also take on a diffusion-like form
while maintaining the angular character of the flux, making them
amenable to solutions with modern diffusion methods.

@@

moments and a set of first-order differential equations:

@@

%%---------------------------------------------------------------------------%%
\section{Spatial Discretization}
\label{sec:spatial_discretization}
To this point, the formulation of the multigroup $SP_N$ equations
presented have discretized the transport equation in angle and energy
but have yet to consider the spatial component of phase space. We
don't explicitly consider it in this document but instead we will
briefly comment on possible means of discretization. Of primary
interest here is the discretization of the diffusive term in
Eq~(\ref{eq:spn_multigroup_system}) and the convective term on the
boundaries in Eq~(\ref{eq:spn_multigroup_bnd}). Many popular choices
for spatial discretization are available here and include finite
difference and finite element formulations. In the Denovo package at
ORNL, Evans has implemented a finite volume scheme for these
equations. Although arbitrary grids could be handled effectively
through a finite element scheme, the rectilinear grid used in Denovo
is easily discretized through the finite volume method in a
conservative form. This conservative form is ideal for the
diffusion-like equations in the domain given by
Eq~(\ref{eq:spn_multigroup_system}) and the effective boundary current
conditions given by Eq~(\ref{eq:spn_multigroup_bnd}) as the neutron
current is balanced from cell-to-cell, continuity of the flux is
preserved across cell/material boundaries, and boundary conditions are
naturally represented through cell-face currents. Furthermore, this
spatial discretization is a natural extension of the balance
principles used to arrive at the general transport equation.

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis of the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Before we can move on to solving the $SP_N$ equations with both
conventional and Monte Carlo methods, we must first understand their
spectral character to ensure the MCSA will apply. As presented in
Chapter~\ref{chap:mc_background}, the spectral radius of the iteration
matrix must be less than unity to ensure convergence and a finite
variance. Therefore, we will compute the spectral radius of the $SP_N$
equations parametrically varying the $SP_N$ order, the $P_N$ order,
the number of energy groups, and upscattering/downscattering between
energy groups.

\subsection{Point Jacobi Spectral Analysis Results}
\label{subsec:spn_analysis_results}

\subsection{Preconditioning Strategy}
\label{subsec:spn_preconditioning}

\subsection{Block Jacobi Spectral Analysis Results}
\label{subsec:spn_analysis_results}
